---
title: "Gender Diversity in R vs Python"
output: html_notebook
---

Guidance on gender of Chinese names: http://blog.tutorming.com/mandarin-chinese-learning-tips/difference-chinese-male-and-female-names-gender

```{r}
library(tidyverse)
library(purrr)
library(jsonlite)
library(rvest)
library(installr)
library(data.table) #for downloading CRAN/RStudio logs

source(file="download_RStudio_CRAN_data.R") 
```

Use the cranlogs api from RStudio to get top package downloads from their CRAN mirror.

```{r get_top_R_packages}
# ----------------------------------------------------------------
#select 5 random days from the last six months
# Read data from RStudio site
RStudio_CRAN_dir <- download_RStudio_CRAN_data(START = Sys.Date()-180,END = Sys.Date(),sample=5)
# read .gz compressed files form local directory
RStudio_CRAN_data <- read_RStudio_CRAN_data(RStudio_CRAN_dir)
 
dim(RStudio_CRAN_data)

# Find the most downloaded packages
r_pkg_list <- most_downloaded_packages(RStudio_CRAN_data,n=100) %>% 
  as_tibble(.name_repair = make.names,c("downloads")) %>% 
  rename(package=X)

save(r_pkg_list,file="data/r_pkg_list.rdata")
```

Source for top 100 Python packages over the last year: https://hugovk.github.io/top-pypi-packages/
Get json (how helpful!).

```{r get_top_python_packages}
py_pkgs_raw<-read_json("https://hugovk.github.io/top-pypi-packages/top-pypi-packages-365-days.json",
                       simplifyVector = TRUE)
python_pkg_list <- py_pkgs_raw$rows[1:100,] %>% 
  as_tibble() %>% 
  rename(package=project,downloads=download_count)
save(python_pkg_list,file="data/python_pkg_list.rdata")

```

Build functions to get contributors to packages and then the real names of those contributors.
Search for relevant repo with just repo name, no user.

https://api.github.com/search/repositories?q=dplyr%20is:name+language:r&sort=stars&order=desc

will return:
      "full_name": "tidyverse/dplyr",

use that to get contributors:

Contributor url: https://api.github.com/repos/tidyverse/dplyr/contributors

will return all contributors with the JSON form:
   "url": "https://api.github.com/users/romainfrancois",

and that will return:
   "name": "Romain François",


```{r get_package_contributors}
library(dplyr)
library(jsonlite)

#this would actually be a list of multiple repos
repo_name <- "pip"
language <- python

get_contrib_names <- function(repo,language="r"){
# we don't know the github username associated with the package to construct a search
# to get the most likely candidate
search_url <- paste0("https://api.github.com/search/repositories?q=",
                     repo_name,
                     ".%20is:name+language:",
                     language,
                     "&sort=stars&order=desc")


# first api call.  would need to would need to loop/map/apply over multiple contributors repo names
all_repos<-jsonlite::read_json(search_url,simplifyVector = TRUE)

regex_str <- paste0(".+/",repo_name,"$")
target_repo<-all_repos$items %>% 
  filter(str_detect(full_name,regex_str)) %>% 
  pull(full_name)
# no error checking if repo is not found

search_url <- paste0("https://api.github.com/repos/",
                     target_repo,
                     "/contributors")
#second api call
contributors <- jsonlite::read_json(search_url,simplifyVector = TRUE)

#third api call. would need to loop/map/apply over multiple contributors
get_name <- function(contrib_url){
  return(jsonlite::read_json(contrib_url)$name)
}

return(get_name(contributors$url[1]))
}

get_contrib_names(repo,language)
```

